Fig2: Candidate genes

### 2020-04-26 analysis/figure-2.Rmd
new fig 2

```{r setup}
library(tidyverse)
library(ggtext)
library(cowplot)
library(ggpubr)
library(knitr)
library(kableExtra)

source("../R/themes.R")
source("../R/wrangledata.R")

@@

table1 <- left_join(candidateDEGS, GOgenesWide) %>%
  select(gene, bldg_lay:n5_n9, parentalcare, parentalbehavior, NCBI) %>%
  mutate(parentalcare = if_else(is.na(parentalcare), " ", "X"),
         parentalbehavior = if_else(is.na(parentalbehavior), " ", "X")) %>%
  rename("Literature" = "parentalcare", "GO" =  "parentalbehavior")
table1$numDEGs <- rowSums(is.na(table1)) # count NAs to know how many are NS
table1 <- table1 %>% 
  mutate(sig = ifelse(numDEGs == 6, "NS", "DEG")) %>% 
  arrange(sig, gene)  %>%  select(-sig, -numDEGs)
table1[is.na(table1)] <- " " # replace NA with blank space so it's pretty
#table1

@@

## Candidate variance stabilized gene expression (vsds)

@@

# `candidatevsd` loaded with source(../R/wrangleata.R)

# significant candidate genes
hypDEGs <- c("CRHBP", "CRHR2", "DRD1", "GNAQ", 
             "HTR2C", "OPRK1", "PGR")
pitDEGs <- c("ESR1", "MEST", "PRL", "PTEN")
gonadDEGs <- c("AVPR1A", "CREBRF", "FOS", "GNAQ", 
               "NR3C1", "OPRM1", "PRLR")

candidateboxplot <- function(whichtissue, whichgenes, whichsex){

  p <- candidatevsd %>%
    filter(tissue %in% whichtissue,
          gene %in% whichgenes,
          sex %in% whichsex) %>%
    mutate(treatment = factor(treatment, levels = charlevels)) %>%
    mutate(treatmentNum = as.numeric(treatment)) %>%
    ggplot(aes(x = treatmentNum, y = counts)) +
    geom_boxplot(aes(fill = treatment, color = sex), outlier.shape = NA) +
    #geom_smooth(aes(color = sex)) +
    geom_jitter(size = 0.25, aes(color = sex)) +
    facet_wrap(~gene, scales = "free_y", nrow = 1) +
    scale_x_continuous(breaks = c(1,2,3,4,5,6,7,8,9),
                       labels = charlevels) +
    scale_fill_manual(values = allcolors) +
    scale_color_manual(values = allcolors) +
    theme_B3() + 
    theme(legend.position = "none",
          strip.text = element_text(face = "italic"),
          axis.text.x = element_blank()) +
    labs(subtitle = whichsex,
         y = "gene expression",
         x = NULL) 
  return(p)
}

a1 <- candidateboxplot("hypothalamus", hypDEGs, "female") + labs(x = NULL, title = "hypothalamus")

b1 <- candidateboxplot("pituitary", pitDEGs, "female") + labs(x = NULL, title = "pituitary")

b2 <- candidateboxplot("pituitary", "PRL", "male") + labs(x = NULL, y = NULL,  title = "pituitary")

b12 <- plot_grid(b1,b2,nrow = 1, labels = c("B", "C"),
          label_size = 8, rel_widths = c(4,1))

c1 <- candidateboxplot("gonad", gonadDEGs, "female") + 
  labs(x = NULL, title = "gonad")

plot_grid(a1, b12, c1, nrow = 3, labels = c("A", " ","D"), 
          label_size = 8)
```

>>>>

```{r setup, message=F}
library(tidyverse)
library(ggtext)
library(cowplot)
library(ggpubr)
library(knitr)
library(kableExtra)
library(corrr)

source("../R/themes.R")
source("../R/functions.R")
source("../R/wrangledata.R")

@@

## table 1 summary candidate genes
table1 <- left_join(candidateDEGS, parentalcaregenes) %>%
  select(gene, bldg_lay:n5_n9, parentalcare, parentalbehavior, NCBI) %>%
  mutate(parentalcare = if_else(is.na(parentalcare), " ", "X"),
         parentalbehavior = if_else(is.na(parentalbehavior), " ", "X")) %>%
  rename("Literature" = "parentalcare", "GO" =  "parentalbehavior")
table1$numDEGs <- rowSums(is.na(table1)) # count NAs to know how many are NS
table1 <- table1 %>% 
  mutate(sig = ifelse(numDEGs == 6, "NS", "DEG")) %>% 
  arrange(sig, gene)  %>%  select(-sig, -numDEGs)
table1[is.na(table1)] <- " " # replace NA with blank space so it's pretty
table1

@@

## Candidate Correlations

```{r}
# load `candidatevsd` with `source("../R/wrangledata.R")`
head(candidatevsd)

curleychampagnegenes <- curleychampagnegenes %>% pull(gene)

hyp1 <- makecorrdf("female", "hypothalamus", curleychampagnegenes)  
pit1 <- makecorrdf("female", "pituitary", curleychampagnegenes)  
gon1 <- makecorrdf("female", "gonad", curleychampagnegenes)

hyp2 <- makecorrdf("male", "hypothalamus", curleychampagnegenes)  
pit2 <- makecorrdf("male", "pituitary", curleychampagnegenes)  
gon2 <- makecorrdf("male", "gonad", curleychampagnegenes)

hyp3 <- makecorrdf(sexlevels, "hypothalamus", curleychampagnegenes)  
pit3 <- makecorrdf(sexlevels, "pituitary", curleychampagnegenes)  
gon3 <- makecorrdf(sexlevels, "gonad", curleychampagnegenes)

candidatevsdwide <- candidatevsd  %>%
    pivot_wider(names_from = gene, values_from = counts) 
FH <- subsetcandidatevsdwide("female", "hypothalamus")
FP <- subsetcandidatevsdwide("female", "pituitary")
FG <- subsetcandidatevsdwide("female", "gonad")
MH <- subsetcandidatevsdwide("male", "hypothalamus")
MP <- subsetcandidatevsdwide("male", "pituitary")
MG <- subsetcandidatevsdwide("male", "gonad") 
```

### Candidate gene functions

```{r}
candidateboxplot <- function(whichtissue, whichgenes, whichsex){

  p <- candidatevsd %>%
    filter(tissue %in% whichtissue,
          gene %in% whichgenes,
          sex %in% whichsex) %>%
    mutate(treatment = factor(treatment, levels = charlevels)) %>%
    mutate(treatmentNum = as.numeric(treatment)) %>%
    ggplot(aes(x = treatmentNum, y = counts)) +
    geom_boxplot(aes(fill = treatment, color = sex), outlier.shape = NA) +
    #geom_smooth(aes(color = sex)) +
    geom_jitter(size = 0.25, aes(color = sex)) +
    facet_wrap(~sex, scales = "free_y", nrow = 2) +
    scale_x_continuous(breaks = c(1,2,3,4,5,6,7,8,9),
                       labels = charlevels) +
    scale_fill_manual(values = allcolors) +
    scale_color_manual(values = allcolors) +
    theme_B3() + 
    theme(legend.position = "none",
          axis.title.y = element_text(face = "italic"),
          axis.text.x = element_blank()) +
    labs(y = whichgenes,
         x = NULL) 
  return(p)
}

scattercorrelations <- function(df, gene1, gene2, mylinecolor, myxlab, myylab){
  p <- ggplot(df, aes(x = gene1, y = gene2)) +
    labs(subtitle = " ", x = myxlab, y = myylab) + 
    scale_color_manual(values = allcolors) +
    geom_smooth(method = "lm",  color = mylinecolor) +
    geom_point(aes(color = treatment))  +
    theme_B3() +  
    theme(axis.title = element_text(face = "italic"), 
            legend.position = "none") +
    stat_cor( size = 2)
  return(p)
}

```

@@

## plot sig candidate genes
a <- candidateboxplot("hypothalamus", "DRD1", sexlevels) + labs(x = NULL, title = "Hypothalamus")

g <- candidateboxplot("pituitary", "PRL", sexlevels) + labs(x = NULL, title = "Pituitary")

m <- candidateboxplot("gonad", "CREBRF", sexlevels) + labs(x = NULL, title = "Gonad") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

c <- plotcorrplot(hyp3, "females and males") + theme(legend.position = "none")  + labs(title = "")
i <- plotcorrplot(pit3, "females and males") + theme(legend.position = "none") + labs(title = "")
o <- plotcorrplot(gon3, "females and males") + theme(legend.position = "bottom")  + labs(title = "")

e <- scattercorrelations(FH, FH$DRD1, FH$HTR2C, "#969696", "DRD1", "HTR2C") + labs(subtitle = "female")+ labs(title = "")
k <- scattercorrelations(FP, FP$AVPR1A, FP$CRHR1, "#969696", "AVPR1A", "CRHR1")  + labs(subtitle = "female")+ labs(title = "")
q <- scattercorrelations(FG, FG$ESR2, FG$CRHBP, "#969696", "ESR2", "CRHBP")   + labs(subtitle = "female")+ labs(title = "")

f <- scattercorrelations(MH, MH$DRD1, MH$HTR2C, "#525252", "DRD1", "HTR2C") + theme(axis.title.y = element_blank()) + labs(subtitle = "male") + labs(title = "")
l <- scattercorrelations(MP, MP$AVPR1A, MP$CRHR1, "#525252", "AVPR1A", "CRHR1") + theme(axis.title.y = element_blank()) + labs(subtitle = "male") + labs(title = "")
r <- scattercorrelations(MG, MG$ESR2, MG$CRHBP, "#525252", "ESR2", "CRHBP")  + theme(axis.title.y = element_blank()) + labs(subtitle = "male") + labs(title = "")

plot_grid(a,c,e,f,g,i,k,l,m,o,q,r, nrow = 3, rel_widths = c(1,1.2,0.8,0.75), rel_heights = c(1,1,1.2),
          labels = "AUTO", label_size =  8)
```
